PoPoolation Pipeline

### 2022-01-29 poolseq.Rmd
add refs
adding figures and examples of output files

---
title: "PoPoolation Pipeline"
subtitle: '*Joanna Griffiths*'
date: "8/12/2021"
bibliography: RCN_site-Nov05.bib
#output:
#  html_document: default
#  pdf_document: default
#  word_document: default
---

@@

@@

## Calculate Fisher's Exact Test
This script will calculate significant allele frequency differences for each SNP position between two samples. Results can be plotted in a Manhattan plot. 
Depending on the experimental method, the `--window-summary-method` flag can be used to determine p-value for windows of SNPs. The default is to multiply the p-value of all SNPs for the window. Alternatively, you can specify how p-values within a window should be summarized. Options include multiple, geometric mean, or median.  The summary of p-values within a window is dependent on the experimental design and how recent selection was expected to occur. For example, the default parameter might be appropriate if your treatment comparisons are geographically separated populations, where selection is expected to have occurred over hundreds of thousands of years. However, this might be inappropriate for short-term experimental evolution studies, such as the example dataset here. In this dataset, copepods were selected in the lab for 15-20 generations, therefore, we expect relatively large linkage blocks that prevent us from identifying the true targets of selection, and therefore I use the geometric mean option.

@@

#You may need to run this on the command line to install twotailed perl Module before running Fisher-test.pl: 
perl -MCPAN -e 'install Text::NSP::Measures::2D::Fisher2::twotailed', see also LSU HPC for details: http://www.hpc.lsu.edu/docs/faq/installation-details.php

@@

## Calculate Exact Allele Frequencies in Samples
```{bash eval=FALSE}
#!/bin/bash

@@

@@

@@

@@

>>>>

---
title: "PoPoolation Pipeline"
subtitle: '*Joanna Griffiths*'
date: "8/12/2021"
bibliography: RCN_site-Nov05.bib
---

@@

Sample of a synchronized file:

```{bash eval = FALSE}
2  2302    N   0:7:0:0:0:0 0:7:0:0:0:0
2  2303    N   0:8:0:0:0:0 0:8:0:0:0:0
2  2304    N   0:0:9:0:0:0 0:0:9:0:0:0
2  2305    N   1:0:9:0:0:0 0:0:9:1:0:0
```

* col1: reference contig
* col2: position within the refernce contig
* col3: reference character
* col4: allele frequencies of population number 1
* col5: allele frequencies of population number 2
* coln: allele frequencies of population number n

The allele frequencies are in the format A:T:C:G:N:del, i.e: count of bases 'A', count of bases 'T',... and deletion count in the end (character '*' in the mpileup)

@@

The following R script can be used to plot Fst results in a Manhattan plot
```{r eval = FALSE}

library("qqman")
library("DataCombine")

fst = read.table("all_PE_w10000_step10000.fst", header = T) 
fst <- fst[c("CHR", "BP", "X1.11", "X1.12", "X2.11", "X2.12", "X3.11", "X3.12", "X4.11", "X4.12", "X5.11", "X5.12", "X6.11", "X6.12", "X7.11", "X7.12", "X8.11", "X8.12", "X9.11", "X9.12", "X10.11", "X10.12", "X11.12")] #rename column headers
fst$ID <- paste(fst$CHR,fst$BP,sep=".")

#tidying file
Replaces <- data.frame(from=c("Chromosome_"), to=c(""))
fst2 <- FindReplace(data = fst, Var = "CHR", replaceData = Replaces,
                       from = "from", to = "to", exact = FALSE)
fst2$CHR <- as.numeric(fst2$CHR)

manhattan(fst2, chr="CHR", bp="BP", snp="ID", p="X1.11", suggestiveline = F, genomewideline = F, logp = F, ylim =c(0,1)) #p here refers to pvalue or in this case the Fst value, substitute with which column comparison you are interested in graphing

```
![Manhattan plot of mean Fst values for 10,000-bp windows of all the selected lines versus the pure Bodega population.](images/Fst_SvsBR_SDmap.png)

## Calculate Fisher's Exact Test
This script will calculate significant allele frequency differences for each SNP position between two samples. Results can be plotted in a Manhattan plot. 
Depending on the experimental method, the `--window-summary-method` flag can be used to determine p-value for windows of SNPs. The default is to multiply the p-value of all SNPs for the window. Alternatively, you can specify how p-values within a window should be summarized. Options include multiply, geometric mean, or median.  The summary of p-values within a window is dependent on the experimental design and how recent selection was expected to occur. For example, the default parameter might be appropriate if your treatment comparisons are geographically separated populations, where selection is expected to have occurred over hundreds of thousands of years. However, this might be inappropriate for short-term experimental evolution studies, such as the example dataset here. In this dataset, copepods were selected in the lab for 15-20 generations, therefore, we expect relatively large linkage blocks that prevent us from identifying the true targets of selection, and therefore I use the geometric mean option.

@@

@@

## Calculate Exact Allele Frequencies in Samples
Exact allele frequencies can be used to determine differences in allele frequencies among populations or before and after a selection experiment, for example. These result could be plotted in a PCA or Manhattan plot.
```{bash eval=FALSE}
#!/bin/bash

@@

```
This script creates two output files having two different extensions:

`_rc`: this file contains the major and minor alleles for every SNP in a concise format
`_pwc`: this file contains the differences in allele frequencies for every pairwise comparision of the populations present in the synchronized file
For details see the man pages of the script
The allele frequency differences can be found in the `_pwc` file, a small sample:

```{bash eval=FALSE}
##chr   pos     rc      allele_count    allele_states   deletion_sum    snp_type        most_variable_allele    diff:1-2
2      4459    N       2       C/T     0       pop     T       0.133
2      9728    N       2       T/C     0       pop     T       0.116
```

The last column contains the obtained differences in allele frequencies for the allele provided in column 8. Note that in this example the last column refers to a pairwise comparison between population 1 vs 2, in case several populations are provided all pairwise comparisons will be appended in additional columns.

@@

```
I then analyzed the results of the CMH test by plotting the geometric mean for each 10,000bp window on a Manhattan plot.

@@

I used nucleotide diversity for creating a PCA plot to compare genetic diversity among my samples and to look for broad patterns across my data and I ran an Adonis test to determine if there were significant differences among groups (in this example, selected vs control lines)
```{r eval=FALSE}

#import each pi file and do some cleanup:
pi_1S <- read.delim("1S.pi", header = F)
colnames(pi_1S) <- c("Chromosome", "window", "num.snps", "frac", "pi_1S")
pi_1S$ID <- paste(pi_1S$Chromosome, pi_1S$window, sep = '_')
pi_1S <- pi_1S[!pi_1S$pi_1S == "na",] #remove rows with na
pi_1S <- pi_1S[,c(6,5)]

pi_1U <- read.delim("1U.pi", header = F)
colnames(pi_1U) <- c("Chromosome", "window", "num.snps", "frac", "pi_1U")
pi_1U$ID <- paste(pi_1U$Chromosome, pi_1U$window, sep = '_')
pi_1U <- pi_1U[!pi_1U$pi_1U == "na",]
pi_1U <- pi_1U[,c(6,5)]

pi_2S <- read.delim("2S.pi", header = F)
colnames(pi_2S) <- c("Chromosome", "window", "num.snps", "frac", "pi_2S")
pi_2S$ID <- paste(pi_2S$Chromosome, pi_2S$window, sep = '_')
pi_2S <- pi_2S[!pi_2S$pi_2S == "na",] #remove rows with na
pi_2S <- pi_2S[,c(6,5)]

pi_2U <- read.delim("2U.pi", header = F)
colnames(pi_2U) <- c("Chromosome", "window", "num.snps", "frac", "pi_2U")
pi_2U$ID <- paste(pi_2U$Chromosome, pi_2U$window, sep = '_')
pi_2U <- pi_2U[!pi_2U$pi_2U == "na",]
pi_2U <- pi_2U[,c(6,5)]

pi_3S <- read.delim("3S.pi", header = F)
colnames(pi_3S) <- c("Chromosome", "window", "num.snps", "frac", "pi_3S")
pi_3S$ID <- paste(pi_3S$Chromosome, pi_3S$window, sep = '_')
pi_3S <- pi_3S[!pi_3S$pi_3S == "na",] #remove rows with na
pi_3S <- pi_3S[,c(6,5)]

pi_4S <- read.delim("4S.pi", header = F)
colnames(pi_4S) <- c("Chromosome", "window", "num.snps", "frac", "pi_4S")
pi_4S$ID <- paste(pi_4S$Chromosome, pi_4S$window, sep = '_')
pi_4S <- pi_4S[!pi_4S$pi_4S == "na",] #remove rows with na
pi_4S <- pi_4S[,c(6,5)]

pi_4U <- read.delim("4U.pi", header = F)
colnames(pi_4U) <- c("Chromosome", "window", "num.snps", "frac", "pi_4U")
pi_4U$ID <- paste(pi_4U$Chromosome, pi_4U$window, sep = '_')
pi_4U <- pi_4U[!pi_4U$pi_4U == "na",]
pi_4U <- pi_4U[,c(6,5)]

pi_5S <- read.delim("5S.pi", header = F)
colnames(pi_5S) <- c("Chromosome", "window", "num.snps", "frac", "pi_5S")
pi_5S$ID <- paste(pi_5S$Chromosome, pi_5S$window, sep = '_')
pi_5S <- pi_5S[!pi_5S$pi_5S == "na",] #remove rows with na
pi_5S <- pi_5S[,c(6,5)]

pi_5U <- read.delim("5U.pi", header = F)
colnames(pi_5U) <- c("Chromosome", "window", "num.snps", "frac", "pi_5U")
pi_5U$ID <- paste(pi_5U$Chromosome, pi_5U$window, sep = '_')
pi_5U <- pi_5U[!pi_5U$pi_5U == "na",]
pi_5U <- pi_5U[,c(6,5)]

pi_6U <- read.delim("6U.pi", header = F)
colnames(pi_6U) <- c("Chromosome", "window", "num.snps", "frac", "pi_6U")
pi_6U$ID <- paste(pi_6U$Chromosome, pi_6U$window, sep = '_')
pi_6U <- pi_6U[!pi_6U$pi_6U == "na",]
pi_6U <- pi_6U[,c(6,5)]

pi_SD <- read.delim("SD.pi", header = F)
colnames(pi_SD) <- c("Chromosome", "window", "num.snps", "frac", "pi_SD")
pi_SD$ID <- paste(pi_SD$Chromosome, pi_SD$window, sep = '_')
pi_SD <- pi_SD[!pi_SD$pi_SD == "na",]
pi_SD <- pi_SD[,c(6,5)]

pi_BR <- read.delim("BR.pi", header = F)
colnames(pi_BR) <- c("Chromosome", "window", "num.snps", "frac", "pi_BR")
pi_BR$ID <- paste(pi_BR$Chromosome, pi_BR$window, sep = '_')
pi_BR <- pi_BR[!pi_BR$pi_BR == "na",]
pi_BR <- pi_BR[,c(6,5)]

#merge all pi results into one file
pi_all1 <- merge(pi_1S, pi_1U, by="ID")
pi_all2 <- merge(pi_all1, pi_2S, by="ID")
pi_all3 <- merge(pi_all2, pi_2U, by="ID")
pi_all4 <- merge(pi_all3, pi_3S, by="ID")
pi_all5 <- merge(pi_all4, pi_4S, by="ID")
pi_all6 <- merge(pi_all5, pi_4U, by="ID")
pi_all7 <- merge(pi_all6, pi_5S, by="ID")
pi_all8 <- merge(pi_all7, pi_5U, by="ID")
pi_all9 <- merge(pi_all8, pi_6U, by="ID")
pi_all10 <- merge(pi_all9, pi_SD, by="ID")
pi_all <- merge(pi_all10, pi_BR, by="ID")

pi_all$pi_1S <- as.numeric(pi_all$pi_1S)
pi_all$pi_1U <- as.numeric(pi_all$pi_1U)
pi_all$pi_2S <- as.numeric(pi_all$pi_2S)
pi_all$pi_2U <- as.numeric(pi_all$pi_2U)
pi_all$pi_3S <- as.numeric(pi_all$pi_3S)
pi_all$pi_4S <- as.numeric(pi_all$pi_4S)
pi_all$pi_4U <- as.numeric(pi_all$pi_4U)
pi_all$pi_5S <- as.numeric(pi_all$pi_5S)
pi_all$pi_5U <- as.numeric(pi_all$pi_5U)
pi_all$pi_6U <- as.numeric(pi_all$pi_6U)
pi_all$pi_BR <- as.numeric(pi_all$pi_BR)
pi_all$pi_SD <- as.numeric(pi_all$pi_SD)

dds.pcoa=pcoa(vegdist(t((pi_all[,2:13])), method="euclidean")/1000)
scores=dds.pcoa$vectors
percent <- dds.pcoa$values$Eigenvalues
percent / sum(percent) #percent for each axes; change in caption below

windows()
plot(scores[,1], scores[,2],
     col=c("red", "blue", "red", "blue", "red", "red", "blue", "red", "blue", "blue", "black", "black"), 
     pch = c(19,19,19,19,19,19,19,19,19,19,1,19),
     xlab = "PC1 (24.33%)", ylab = "PC2 (21.24%)")

treat_df <- data.frame(
  ID = c("1S_pi", "1U_pi", "2S_pi", "2U_pi", "3S_pi", "4S_pi", "4U_pi", "5S_pi", "5U_pi","6U_pi"),
  line = c("S", "U", "S", "U", "S", "S", "U", "S", "U", "U")
)

line <- treat_df$line

adonis2(t(pi_all[,2:11])~line, data = treat_df, permutations = 1000000, method = "manhattan")

```
![Principle coordinate analysis of allele frequencies for selected lines (red), control lines (blue), the pure Bodega population (white), and the pure San Diego population (black).](images/pi_PCA.png)

@@

Griffiths JS, Kawji Y, Kelly MW. 2020. An experimental test of adaptive introgression in locally adapted populations of splash pool copepods. Mol Biol and Evol. 38(4):1306-1316. https://doi.org/10.1093/molbev/msaa289

Schlötterer C, Tobler R, Kofler R, Nolte V. 2014 Sequencing pools of individuals-mining genome-wide polymorphism data without big funding. Nat. Rev. Genet. 15, 749–763. (doi:10.1038/nrg3803)
